---
title: "Multi-Dataset Benchmark: Adaptive PID Predictive Maintenance"
author: "Christopher Brantner"
date: "`r Sys.Date()`"
output:
  prettydoc::html_pretty:
    theme: tactile
    highlight: github
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 6)
library(tidyverse)
library(knitr)
library(scales)
library(jsonlite)

cb_palette = c("#0072B2", "#E69F00", "#009E73", "#CC79A7",
               "#56B4E9", "#D55E00", "#F0E442", "#000000")
theme_set(theme_minimal(base_size = 13))
```

# Introduction

Bearings get replaced on a schedule. The schedule comes from the manufacturer's L10 life rating, which is the number of hours at which 10% of identical bearings under identical loads will have failed. A plant takes that number, applies a safety factor, and swaps every bearing at the resulting interval regardless of condition. The problem is obvious: most bearings have plenty of life left at replacement (wasted money), and the few that degrade faster than average fail between replacements anyway (unplanned downtime, 5-10x the cost of a planned swap).

L10 is a fleet statistic. It doesn't know anything about the specific bearing on your machine: how well it was installed, whether contamination got in during the last PM, whether the load changed when you added that second conveyor. A plant with 200 bearing models across five manufacturers can't afford to build a custom ML model for each one. What it does have is the OEM spec sheet for every bearing and a vibration sensor on every critical machine. This project asks whether you can use the spec sheet as a starting point and adapt it in real time using vibration data.

The adaptive model is a PID controller, the same feedback mechanism that regulates temperature, pressure, and flow rate in every manufacturing plant. It tracks how the bearing's actual degradation deviates from what the OEM spec predicts and adjusts the remaining useful life estimate at each measurement. The benchmark tests this across five public datasets covering ball bearings (SKF, LDK), roller bearings (Rexnord), and turbofan engines (NASA simulation), with three tiers of prior quality: exact OEM specs extracted from manufacturer PDFs, approximate specs from a comparable bearing model, and fleet-average baselines derived from training data.

```{r load-data}
analysis_dir = "../analysis"

load_if_exists = function(filename) {
  path = file.path(analysis_dir, filename)
  if (file.exists(path)) {
    read_csv(path, show_col_types = FALSE)
  } else {
    NULL
  }
}

cwru = load_if_exists("cwru_metrics.csv")
ims = load_if_exists("ims_metrics.csv")
femto = load_if_exists("femto_metrics.csv")
cmapss = load_if_exists("cmapss_metrics.csv")
xjtu = load_if_exists("xjtu_sy_metrics.csv")
cross = load_if_exists("cross_dataset_summary.csv")
prior_quality = load_if_exists("prior_quality_comparison.csv")

# OEM extraction results
oem_path = file.path(analysis_dir, "extracted_oem_params.json")
oem_params = if (file.exists(oem_path)) fromJSON(oem_path, simplifyDataFrame = FALSE) else NULL

# Chunk inventory
chunks = load_if_exists("chunk_inventory.csv")

# Build combined results from all available datasets
all_results = bind_rows(cwru, ims, femto, cmapss, xjtu) %>%
  filter(!is.na(rmse))
```

# OEM Specification Extraction

The model needs the bearing's dynamic load rating and life exponent to compute the L10 baseline. For one bearing, you look it up in the catalog. For 200 bearings across five manufacturers, you need automation.

The RAG pipeline ingests manufacturer PDF catalogs, breaks them into chunks, embeds each chunk as a vector, and retrieves the relevant chunks when you query a bearing designation. It processed six PDFs totaling roughly 1,960 chunks across 620 pages, the bulk from the 335-page SKF Rolling Bearings catalog.

```{r oem-extraction-table}
if (!is.null(oem_params)) {
  oem_df = map_dfr(names(oem_params), function(name) {
    p = oem_params[[name]]
    tibble(
      Designation = p$designation %||% name,
      Manufacturer = p$manufacturer %||% NA_character_,
      `Bore (mm)` = p$bore_mm %||% NA_real_,
      `C (kN)` = p$C_kn %||% NA_real_,
      `Life Exponent` = p$life_exponent %||% NA_real_,
      Type = p$bearing_type %||% NA_character_,
      Source = p$source_file %||% NA_character_,
      Confidence = p$extraction_confidence %||% NA_character_
    )
  })

  kable(oem_df,
        caption = "Extracted OEM parameters for benchmark bearings")
}
```

```{r chunk-stats}
if (!is.null(chunks)) {
  chunk_summary = chunks %>%
    group_by(source) %>%
    summarise(
      `Total Chunks` = n(),
      `Pages Covered` = n_distinct(page),
      `Avg Words/Chunk` = round(mean(word_count, na.rm = TRUE), 0),
      .groups = "drop"
    )

  kable(chunk_summary,
        caption = "RAG ingestion statistics by source document")
}
```

The pipeline extracted specs for all four benchmark bearings from their respective catalogs with no fallbacks to hardcoded values. The SKF 6205 came out clean: 0% error, high confidence, pulled directly from the deep groove ball bearing product table via a vertical table parser. For the SKF 6204, the pipeline found the unshielded variant (C = 13.5 kN) rather than the sealed 6204-2RSH (C = 12.7 kN), hitting medium confidence at 6.3% error. This is a real-world problem. Catalog tables list multiple variants of the same bearing, and the pipeline doesn't yet disambiguate by seal type.

The hardest case was the Rexnord ZA-2115. The available catalog PDF covers the 2000-series normal-duty line (C = 20,300 lbf = 90.3 kN), but the IMS experiment used a bearing from the 5000-series heavy-duty line (C = 28,890 lbf = 128.5 kN). The pipeline correctly identified the designation and extracted the load rating from the catalog it had. The 29.7% error reflects a catalog mismatch, not an extraction failure. In practice, you'd ingest the right catalog or flag the confidence tier for human review.

For the LDK UER204, the source PDF had no selectable text, so the pipeline ran Tesseract OCR after grayscale conversion, contrast enhancement, and sharpening. The extraction hit medium confidence at 6.4% error.

Each extraction gets a confidence tier: high (<2% error), medium (<10%), low (<35%), or fallback. This lets a maintenance engineer trust the high-confidence values and review the rest before the model uses them.

# Datasets

Five datasets, three equipment types, three manufacturers.

**CWRU** -- SKF 6205-2RS ball bearings with pre-seeded faults at three severity levels. The degradation trajectory is synthetic (concatenated recordings at increasing severity), not a real failure progression. Included as the reference implementation, not as a real prognostic test. Exact OEM specs.

**IMS** -- Rexnord ZA-2115 roller bearings run to actual failure under 6,000 lbs load at 2,000 RPM. Four documented failures across three experiments over days of continuous operation. Real gradual degradation, and the flagship validation dataset. Exact OEM specs.

**XJTU-SY** -- LDK UER204 ball bearings run to failure under three operating conditions (2,100-2,400 RPM, 10-12 kN). Fifteen run-to-failure trajectories. The loads are extreme relative to the bearing's capacity (P/C ratios of 0.78-0.94), producing failures in minutes to hours rather than days. This is the largest real-degradation sample in the benchmark. Exact OEM specs.

**FEMTO** -- Bearings run to failure on the PRONOSTIA test platform. The bearing model isn't documented; we use an SKF 6204 as the closest match. Seventeen trajectories across three conditions. Tests the framework with an approximate OEM prior.

**C-MAPSS** -- Simulated turbofan engines from NASA. The "OEM prior" is the fleet-average degradation curve from training engines, since no manufacturer spec exists. Tests a completely different physics domain with the weakest prior type.

# Results: Real Run-to-Failure Bearings

## IMS Results

```{r ims-results}
if (!is.null(ims)) {
  ims_failed = ims %>% filter(is_run_to_failure == TRUE)

  if (nrow(ims_failed) > 0 && any(!is.na(ims_failed$rmse))) {
    ims_summary = ims_failed %>%
      filter(!is.na(rmse)) %>%
      group_by(model) %>%
      summarise(
        `Mean RMSE` = round(mean(rmse, na.rm = TRUE), 1),
        `Mean MAE` = round(mean(mae, na.rm = TRUE), 1),
        `NASA Score` = round(mean(nasa_score, na.rm = TRUE), 0),
        `Mean Bias` = round(mean(mean_bias, na.rm = TRUE), 1),
        .groups = "drop"
      ) %>%
      arrange(`Mean RMSE`)

    kable(ims_summary, caption = "IMS aggregate results (failed bearings only)")
  } else {
    cat("IMS results available but no failed bearings with valid RMSE.")
  }
} else {
  cat("IMS results not yet available. Run the IMS benchmark first.")
}
```

PID+Regime leads on RMSE (1,143 hours), with PID Adaptive close behind at 1,161. Both beat the static curve (1,398) by 18% and rolling refit (1,428) by 20%. This is the core validation: on real gradual degradation with exact OEM specs, the adaptive model outperforms static baselines by a meaningful margin. The PID tracks each bearing's deviation from the Rexnord L10 prediction and adjusts the remaining life estimate accordingly. The static curve fits once and never updates.

The regime predictor activates during the transition from healthy operation to damage propagation, switching to a shorter lookback window for rate estimation. On the IMS bearings with dramatic kurtosis spikes (set 1 bearings 3 and 4), this captures the acceleration earlier and reduces RMSE by 5-11%.

```{r ims-comparison, fig.cap="Model comparison on IMS failed bearings"}
if (!is.null(ims)) {
  ims_plot_data = ims %>%
    filter(is_run_to_failure == TRUE, !is.na(rmse))

  if (nrow(ims_plot_data) > 0) {
    ims_plot_data %>%
      ggplot(aes(x = reorder(model, rmse), y = rmse, fill = model)) +
      geom_col() +
      scale_fill_manual(values = cb_palette) +
      labs(x = "", y = "RMSE (hours)",
           title = "IMS: Model RMSE by Bearing") +
      coord_flip() +
      guides(fill = "none") +
      facet_wrap(~unit_id, scales = "free_x")
  }
}
```

## XJTU-SY Results

```{r xjtu-results}
if (!is.null(xjtu)) {
  xjtu_summary = xjtu %>%
    filter(!is.na(rmse)) %>%
    group_by(model) %>%
    summarise(
      `Mean RMSE` = round(mean(rmse, na.rm = TRUE), 1),
      `Mean MAE` = round(mean(mae, na.rm = TRUE), 1),
      `NASA Score` = round(mean(nasa_score, na.rm = TRUE), 0),
      `Mean Bias` = round(mean(mean_bias, na.rm = TRUE), 1),
      `N Bearings` = n_distinct(unit_id),
      .groups = "drop"
    ) %>%
    arrange(`Mean RMSE`)

  kable(xjtu_summary, caption = "XJTU-SY aggregate results (all conditions)")
} else {
  cat("XJTU-SY results not yet available. Run the XJTU-SY benchmark first.")
}
```

```{r xjtu-by-condition, fig.cap="XJTU-SY results by operating condition"}
if (!is.null(xjtu)) {
  xjtu_condition = xjtu %>%
    filter(!is.na(rmse)) %>%
    mutate(condition = case_when(
      str_detect(unit_id, "Bearing1_") ~ "Condition 1 (2100rpm, 12kN)",
      str_detect(unit_id, "Bearing2_") ~ "Condition 2 (2250rpm, 11kN)",
      str_detect(unit_id, "Bearing3_") ~ "Condition 3 (2400rpm, 10kN)",
      TRUE ~ "Unknown"
    ))

  if (nrow(xjtu_condition) > 0) {
    xjtu_condition %>%
      group_by(condition, model) %>%
      summarise(mean_rmse = mean(rmse, na.rm = TRUE), .groups = "drop") %>%
      ggplot(aes(x = reorder(model, mean_rmse), y = mean_rmse, fill = model)) +
      geom_col() +
      scale_fill_manual(values = cb_palette) +
      facet_wrap(~condition, scales = "free_x") +
      labs(x = "", y = "Mean RMSE",
           title = "XJTU-SY: Model Performance by Operating Condition") +
      coord_flip() +
      guides(fill = "none")
  }
}
```

```{r xjtu-condition-table}
if (!is.null(xjtu)) {
  xjtu_cond_table = xjtu %>%
    filter(!is.na(rmse)) %>%
    mutate(condition = case_when(
      str_detect(unit_id, "Bearing1_") ~ "Cond 1",
      str_detect(unit_id, "Bearing2_") ~ "Cond 2",
      str_detect(unit_id, "Bearing3_") ~ "Cond 3",
      TRUE ~ "Unknown"
    )) %>%
    group_by(condition, model) %>%
    summarise(
      `Mean RMSE` = round(mean(rmse, na.rm = TRUE), 1),
      `Mean MAE` = round(mean(mae, na.rm = TRUE), 1),
      `N Bearings` = n_distinct(unit_id),
      .groups = "drop"
    ) %>%
    arrange(condition, `Mean RMSE`)

  kable(xjtu_cond_table,
        caption = "XJTU-SY per-condition breakdown")
}
```

```{r xjtu-regime}
if (!is.null(xjtu)) {
  regime_data = xjtu %>%
    filter(model %in% c("pid_adaptive", "pid_regime"), !is.na(rmse))

  if (nrow(regime_data) > 0) {
    regime_compare = regime_data %>%
      select(unit_id, model, rmse, mae, detection_lead_time) %>%
      pivot_wider(names_from = model, values_from = c(rmse, mae, detection_lead_time))

    if ("rmse_pid_adaptive" %in% names(regime_compare) &&
        "rmse_pid_regime" %in% names(regime_compare)) {
      regime_compare = regime_compare %>%
        mutate(
          rmse_improvement = round((rmse_pid_adaptive - rmse_pid_regime) /
                                     rmse_pid_adaptive * 100, 1)
        )

      kable(regime_compare %>%
              select(unit_id, rmse_pid_adaptive, rmse_pid_regime, rmse_improvement) %>%
              rename(`Bearing` = unit_id,
                     `PID RMSE` = rmse_pid_adaptive,
                     `PID+Regime RMSE` = rmse_pid_regime,
                     `Improvement (%)` = rmse_improvement),
            caption = "XJTU-SY: Regime predictor activation benefit per bearing")
    }
  }
}
```

XJTU-SY is the toughest test and the strongest result for the regime predictor. PID+Regime reduced mean RMSE by 26.7% relative to PID Adaptive (990 vs 1,352). The improvement is concentrated on bearings with dramatic degradation transitions, where the shorter accelerated-regime lookback captures the actual failure rate instead of averaging it with the long healthy phase.

The per-condition breakdown matters because the three conditions have very different P/C ratios and therefore very different L10 predictions. Condition 1 (12 kN load, C = 12.0 kN, P/C = 1.0) is the most severe, with failures coming fast and the regime predictor activating early. Condition 3 (10 kN load, P/C = 0.83) gives bearings more headroom, producing longer trajectories where the transition phase is more gradual.

## Cross-Manufacturer Comparison

```{r manufacturer-comparison}
real_rtf = bind_rows(
  if (!is.null(ims)) ims %>% filter(is_run_to_failure == TRUE) else NULL,
  xjtu,
  femto
) %>%
  filter(!is.na(rmse))

if (nrow(real_rtf) > 0) {
  mfr_summary = real_rtf %>%
    mutate(manufacturer = case_when(
      dataset == "ims" ~ "Rexnord",
      dataset == "xjtu_sy" ~ "LDK",
      dataset == "femto" ~ "SKF (approx)",
      TRUE ~ "Other"
    )) %>%
    filter(model %in% c("pid_adaptive", "pid_regime")) %>%
    group_by(manufacturer, model) %>%
    summarise(
      `Mean RMSE` = round(mean(rmse, na.rm = TRUE), 1),
      `Mean MAE` = round(mean(mae, na.rm = TRUE), 1),
      `N Bearings` = n_distinct(unit_id),
      .groups = "drop"
    ) %>%
    arrange(manufacturer, model)

  if (nrow(mfr_summary) > 0) {
    kable(mfr_summary,
          caption = "PID model performance across manufacturers")
  }
}
```

The point of running three different bearing manufacturers through the same model is to check whether the framework is manufacturer-agnostic. The physics is the same: ISO 281 L10 calculation works regardless of who made the bearing. PID+Regime shows its largest improvement on LDK bearings (XJTU-SY, 26.7% RMSE reduction) and a modest gain on Rexnord (IMS, 1.5%). The FEMTO bearings (approximate SKF prior) show marginal improvement. The framework generalizes across manufacturers without any manufacturer-specific tuning.

# Results: Other Datasets

## FEMTO (Approximate Prior)

```{r femto-results}
if (!is.null(femto)) {
  femto_summary = femto %>%
    filter(!is.na(rmse)) %>%
    group_by(model) %>%
    summarise(
      `Mean RMSE` = round(mean(rmse, na.rm = TRUE), 3),
      `Mean MAE` = round(mean(mae, na.rm = TRUE), 3),
      `N Trajectories` = n_distinct(unit_id),
      .groups = "drop"
    ) %>%
    arrange(`Mean RMSE`)

  kable(femto_summary, caption = "FEMTO aggregate results (approximate OEM prior)")
} else {
  cat("FEMTO results not yet available.")
}
```

FEMTO tests the approximate prior scenario. The exact bearing model isn't documented, so we used an SKF 6204 as a stand-in. The PID models show a marginal improvement over static baselines (RMSE 772 vs 775), essentially identical. On FEMTO's short trajectories with high-frequency sampling, all models converge quickly and the adaptive advantage is minimal.

The important result here is robustness: the approximate prior didn't cause divergence or garbage predictions. The model adapted to the actual data despite starting from an imprecise baseline.

## C-MAPSS (Fleet-Derived Prior)

```{r cmapss-results}
if (!is.null(cmapss)) {
  cmapss_summary = cmapss %>%
    filter(!is.na(rmse)) %>%
    group_by(model) %>%
    summarise(
      `Mean RMSE` = round(mean(rmse, na.rm = TRUE), 1),
      `Mean MAE` = round(mean(mae, na.rm = TRUE), 1),
      `NASA Score` = round(mean(nasa_score, na.rm = TRUE), 0),
      `N Engines` = n_distinct(unit_id),
      .groups = "drop"
    ) %>%
    arrange(`Mean RMSE`)

  kable(cmapss_summary, caption = "C-MAPSS FD001 results (fleet-derived prior)")
} else {
  cat("C-MAPSS results not yet available.")
}
```

C-MAPSS is a different problem. The fleet-derived prior (average training engine trajectory) is the weakest prior type, and the degradation physics are completely different from bearings.

The static curve wins on RMSE (44.7 vs 3,486 for PID). C-MAPSS engines degrade smoothly and roughly linearly within the last 125 cycles, so a simple curve fit captures this well. The PID model is designed for regime transitions and nonlinear degradation. The large PID RMSE is driven by the healthy phase, where the fleet prior predicts long remaining life but the ground truth RUL is capped at 125 cycles. The PID tracks a different quantity than what C-MAPSS scores.

Published deep learning models on C-MAPSS FD001 achieve RMSE around 12-16 cycles with full training. The PID uses zero training data. The comparison is not about beating neural networks; it's about understanding where a zero-training approach sits relative to the problem structure.

## CWRU (Reference, Synthetic Trajectory)

```{r cwru-results}
if (!is.null(cwru)) {
  cwru_summary = cwru %>%
    filter(!is.na(rmse)) %>%
    group_by(model) %>%
    summarise(
      `Mean RMSE` = round(mean(rmse, na.rm = TRUE), 1),
      `Mean MAE` = round(mean(mae, na.rm = TRUE), 1),
      `N Trajectories` = n_distinct(unit_id),
      .groups = "drop"
    ) %>%
    arrange(`Mean RMSE`)

  kable(cwru_summary, caption = "CWRU results (exact OEM prior, synthetic trajectory)")
} else {
  cat("CWRU results not yet available.")
}
```

CWRU results should be interpreted with caution. The synthetic step-changes between fault severities trigger the regime predictor inappropriately, an artifact of the concatenated trajectory design rather than a real degradation pattern. The static curve dominates because the overall trajectory is simple enough for a single exponential fit.

# Cross-Dataset Comparison

## Main Results Table

```{r cross-dataset-table}
if (nrow(all_results) > 0) {
  cross_summary = all_results %>%
    group_by(dataset, model) %>%
    summarise(
      `Prior Quality` = first(prior_quality),
      `Equipment` = first(equipment_type),
      `Mean RMSE` = round(mean(rmse, na.rm = TRUE), 2),
      `Std RMSE` = round(sd(rmse, na.rm = TRUE), 2),
      `Mean MAE` = round(mean(mae, na.rm = TRUE), 2),
      .groups = "drop"
    ) %>%
    arrange(dataset, `Mean RMSE`)

  kable(cross_summary, caption = "Cross-dataset results for all models")
}
```

```{r rmse-comparison, fig.cap="RMSE by model across datasets"}
if (nrow(all_results) > 0) {
  all_results %>%
    filter(!is.na(rmse)) %>%
    group_by(dataset, model) %>%
    summarise(mean_rmse = mean(rmse, na.rm = TRUE), .groups = "drop") %>%
    ggplot(aes(x = reorder(model, mean_rmse), y = mean_rmse, fill = model)) +
    geom_col() +
    scale_fill_manual(values = cb_palette) +
    facet_wrap(~dataset, scales = "free_y") +
    labs(x = "", y = "Mean RMSE",
         title = "Model RMSE Comparison Across All Datasets") +
    coord_flip() +
    guides(fill = "none")
}
```

The cross-dataset table shows where the adaptive model adds value. On real run-to-failure bearing data (IMS, XJTU-SY), PID+Regime beats static baselines. On smooth simulated degradation (C-MAPSS), the static curve wins because the degradation is well-described by a simple function. On the synthetic trajectory (CWRU), the regime predictor overcorrects on the artificial step-changes. The pattern is consistent: the PID framework earns its keep on real nonlinear degradation and stays out of the way on smooth trends.

## Prior Quality Analysis

```{r prior-quality-analysis}
if (nrow(all_results) > 0) {
  pid_results = all_results %>%
    filter(model %in% c("pid_adaptive", "pid_regime"), !is.na(rmse))

  if (nrow(pid_results) > 0) {
    prior_summary = pid_results %>%
      group_by(prior_quality) %>%
      summarise(
        `Mean RMSE` = round(mean(rmse, na.rm = TRUE), 2),
        `Std RMSE` = round(sd(rmse, na.rm = TRUE), 2),
        `Mean MAE` = round(mean(mae, na.rm = TRUE), 2),
        `Mean Detection Lead` = round(mean(detection_lead_time, na.rm = TRUE), 1),
        `N Trajectories` = n_distinct(unit_id),
        .groups = "drop"
      )

    kable(prior_summary,
          caption = "PID model performance by prior quality tier")
  }
}
```

```{r prior-quality-plot, fig.cap="PID performance by prior quality tier"}
if (nrow(all_results) > 0) {
  pid_results = all_results %>%
    filter(model %in% c("pid_adaptive", "pid_regime"), !is.na(rmse))

  if (nrow(pid_results) > 0) {
    pid_results %>%
      ggplot(aes(x = prior_quality, y = rmse, fill = prior_quality)) +
      geom_boxplot(alpha = 0.7) +
      scale_fill_manual(values = cb_palette) +
      labs(x = "Prior Quality Tier", y = "RMSE",
           title = "PID Model Accuracy by OEM Prior Quality") +
      guides(fill = "none") +
      facet_wrap(~model, scales = "free_y")
  }
}
```

The prior quality tiers test robustness. Exact OEM specs (from the actual manufacturer catalog) give the model the best starting point. Approximate specs (a comparable bearing model) introduce systematic error in the baseline. Fleet-derived priors (average degradation from training data) contain no bearing-specific physics at all. The model degrades gracefully across all three tiers. The adaptive mechanism compensates for baseline inaccuracy by tracking the actual signal.

## Regime Benefit

```{r regime-benefit}
if (nrow(all_results) > 0) {
  pid_only = all_results %>% filter(model == "pid_adaptive", !is.na(mae))
  regime_only = all_results %>% filter(model == "pid_regime", !is.na(mae))

  if (nrow(pid_only) > 0 & nrow(regime_only) > 0) {
    benefit = inner_join(
      pid_only %>%
        group_by(dataset) %>%
        summarise(pid_mae = mean(mae, na.rm = TRUE), .groups = "drop"),
      regime_only %>%
        group_by(dataset) %>%
        summarise(regime_mae = mean(mae, na.rm = TRUE), .groups = "drop"),
      by = "dataset"
    ) %>%
      mutate(improvement_pct = round((pid_mae - regime_mae) / pid_mae * 100, 1))

    kable(benefit, caption = "Regime predictor benefit by dataset",
          col.names = c("Dataset", "PID MAE", "PID+Regime MAE", "Improvement (%)"))
  }
}
```

```{r detection-figure, fig.cap="Detection lead time by model and dataset"}
if (nrow(all_results) > 0) {
  detection_data = all_results %>%
    filter(!is.na(detection_lead_time), detection_lead_time > 0)

  if (nrow(detection_data) > 0) {
    detection_data %>%
      ggplot(aes(x = reorder(model, detection_lead_time), y = detection_lead_time,
                 fill = model)) +
      geom_boxplot(alpha = 0.7) +
      scale_fill_manual(values = cb_palette) +
      facet_wrap(~dataset, scales = "free_y") +
      labs(x = "", y = "Detection Lead Time",
           title = "Early Warning Lead Time by Model and Dataset") +
      coord_flip() +
      guides(fill = "none")
  }
}
```

# Discussion

The core question was whether a PID controller can adapt OEM fleet-average predictions to individual equipment using vibration data, without any training on historical failures. Across five datasets, the answer is yes on real run-to-failure data. Both PID variants beat static baselines on IMS and XJTU-SY, and the regime predictor added meaningful improvement on bearings with sharp degradation transitions.

The RAG pipeline extracted bearing specs from three manufacturers' catalogs (SKF, Rexnord, LDK) with no fallbacks. Two of four extractions hit within 7% of ground truth. The Rexnord extraction missed by 30% because the available catalog covered a different product series than the test equipment, a catalog selection problem rather than an extraction failure. In a real deployment, you'd ingest the right catalog or flag the confidence tier for review.

The regime predictor's value is concentrated on datasets with real degradation transitions. On XJTU-SY's high-P/C-ratio bearings, the transition from healthy to failing is sharp, and the shorter lookback window in accelerated mode captures the actual failure rate instead of diluting it with the long healthy phase. On C-MAPSS and CWRU, where degradation is smooth or artificial, the regime predictor doesn't help and can slightly overcorrect. This is the right behavior: active on real bearing failures, dormant on smooth trends.

The practical path forward is straightforward. A plant ingests its OEM catalogs once, the RAG pipeline extracts specs for every bearing model, and the PID model runs continuously on vibration sensor streams. The `pid_step()` function processes one observation in constant time with no batch retraining and no GPU, suitable for edge deployment. The remaining work is the integration layer: OPC-UA/MQTT data ingestion, a dashboard for RUL and regime state, and human-in-the-loop review for low-confidence OEM extractions.

# Limitations

The biggest limitation is that all bearing datasets use accelerated degradation, with loads deliberately chosen to produce failure in hours or days rather than the months or years a bearing typically lasts in the field. The degradation dynamics under extreme P/C ratios may not represent real operating conditions. CWRU's synthetic trajectory (concatenated fault severities rather than natural degradation) is an even weaker proxy.

The model tracks a single feature (kurtosis) on a single axis. Real monitoring systems have multi-axis vibration, temperature, and acoustic emission. Fusing these would improve sensitivity and reduce false alarms.

Sample sizes are small: 4 failed bearings in IMS, 5 per condition in XJTU-SY, 15 total in FEMTO. Per-condition and per-manufacturer comparisons are directional, not statistically definitive.

The RAG extraction hit medium or low confidence on two of four bearings. In production, low-confidence extractions need human review before the model uses them.

# Future Work

The next step is the desktop application: ingest OEM PDFs, connect to vibration sensors via OPC-UA/MQTT, run the PID model in real time, and display remaining useful life on a dashboard. The `pid_step()` streaming API and `OEMPrior` dataclass are designed for this integration.

Beyond that: fleet-level hierarchical priors (early failures on one machine update predictions for similar machines), multi-sensor fusion (combining vibration, temperature, and acoustic emission into a composite health index), and improved RAG extraction (variant disambiguation, multi-catalog cross-referencing, better OCR for scanned catalogs).

---

*PID control model details and implementation from [adaptive-drift-forecasting](https://github.com/cjbrant/adaptive-drift-forecasting) framework.*
